Spatiotemporal models of species biomass with time-varying depth and trawl footprint, and optional substrate variables

### 2019-09-26 analysis/VOCC/2-biomass-depth-only-model.Rmd
update model choice and style

Species
```{r}
species <- params$species

# # Species run so far...
# species <- "Arrowtooth Flounder" # temp good, not DO, imm moving deeper?

# species <- "Petrale Sole"
# species <- "English Sole"
# species <- "Dover Sole"
# species <- "Southern Rock Sole"

# species <- "Walleye Pollock" # survey 4 or 16 might work
# species <- "Pacific Cod"

## Sebastes

## TRY BOTH - not schooling
 # species <- "Yelloweye Rockfish" # beautiful curves, with depth change ***
 #  species <- "Bocaccio" # good curves, no depth change ... most survey-specific models fail to find minimum ***
#  species <- "Sharpchin Rockfish" # beautiful curves, imm depth change ***
#  species <- "Greenstriped Rockfish" # beautiful curves for adults, no depth change ***

# ## TRY BOTH - schooling
#  species <- "Pacific Ocean Perch" # beautiful curves, no depth change, schooling ***
#  species <- "Silvergray Rockfish" # beautiful curves, no depth change, schooling ***

## TRY JUST TEMPERATURE 
# species <- "Rougheye/Blackspotted Rockfish Complex" # good temp, not DO
# species <- "Redbanded Rockfish" # good temp, not DO, maybe moving shallower
 species <- "Splitnose Rockfish" # good temp, survey 4 DO ok

# species <- "Yellowtail Rockfish" #  good temp for survey 4 only, no DO, schooling
# species <- "Quillback Rockfish" # like warm = overall a bit flat... territorial! 
# species <- "Canary Rockfish" # schooling, winter-birthing, and depth change!, but curves at edges... adult1 looks ok?
#  species <- "Widow Rockfish" # schooling, and depth change, but curves at edges

# species <- "Longspine Thornyhead" # too deep to produce curves
# species <- "Shortspine Thornyhead"
# species <- "Sablefish" # temp might work, no DO, no consistant depth change
# species <- "Lingcod" # curves ok, overall a bit flat
# # species <- "Pacific Hake" # not useful

# species <- "North Pacific Spiny Dogfish" # too flat, but with depth change! note: pooled maturity 
# species <- "Longnose Skate" # do too flat, temp from 4 might be useful, seems to go deeper?
# species <- "Big Skate"
# species <- "Sandpaper Skate"
# species <- "Brown Cat Shark"
# species <- "Spotted Ratfish" # curves not useful because to abundant in shallow
```

@@

knitr::opts_chunk$set(fig.width=11, fig.height=8.5, 
                      fig.path=paste0("figs/", spp, "/"),
                      echo=FALSE, warning=FALSE, message=FALSE)
```

@@

  # scale predictors before filtering to ensure mean and SD are global
  data <- data %>% mutate(raw_depth = depth, depth = log(raw_depth)) 
  data <- gfranges::scale_predictors(data, #predictors = c(quo(depth)))
        predictors = c(quo(depth), quo(mixed), quo(muddy), quo(sandy), quo(rocky), quo(any_rock)))
  data <- data %>% mutate(depth = raw_depth) %>% filter(ssid %in% model_ssid)
```

@@

Run climate independent sdmTMB model
```{r}

  if (any(names(data) == "adult_density")) {

    adult_formula <- as.formula(paste(
      "adult_density ~ 0 + as.factor(year)", covariates, ""
      ))

    starttime1 <- Sys.time()
    adult_biomass <- sdmTMB::sdmTMB(data,
      adult_formula,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 , #+ trawled
      time = "year", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = priors,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

    saveRDS(adult_biomass, 
      file = paste0("data/", spp, 
        "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
        ))

@@

    saveRDS(imm_biomass, 
     file = paste0("data/", spp, 
       "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
       ))

  } else {

    starttime3 <- Sys.time()
    dens_formula <- as.formula(paste("density ~ 0 + as.factor(year)", covariates, ""))

    total_biomass <- sdmTMB::sdmTMB(data,
      dens_formula,
      time_varying = ~ 0 + depth_scaled + depth_scaled2,# + trawled
      time = "year", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = priors,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime3 <- Sys.time()
    time3 <- round(starttime3 - endtime3)

    saveRDS(total_biomass, file = paste0("data/", spp, 
      "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
      ))
  }
```

@@

  bath <- readRDS("data/bathymetry-data")
  #dat <- bath$data %>% scale_survey_predictors()
  dat <- bath$data %>% mutate(raw_depth = depth, depth = log(raw_depth)) 
  dat <- gfranges::scale_predictors(dat, predictors = c(quo(depth)))

@@

  newcov <- readRDS("data/new_covariates.rds") %>% dplyr::select(X, Y, trawled, muddy, rocky, mixed, any_rock)
  nd_all <- left_join(nd_all, newcov, by=c("X", "Y"))

@@

```{r}
#covs <- gsub("\\+", "-", covariates)
fish <- readRDS(paste0("raw/bio-data-", spp, ""))

@@

if (length(unique(fish$maturity_code)) > 2) {

    nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
    adult_biomass <- readRDS(paste0("data/", spp, 
      "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    ))

@@

    adult_predictions <- predict(adult_biomass, newdata = nd)
    saveRDS(adult_predictions, file = paste0("data/", spp, 
      "/predictions-", spp, covs, "-", ssid_string, "-mature-biomass-prior-", priors, ".rds"
    ))

    imm_biomass <- readRDS(paste0("data/", spp, 
      "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    ))
    imm_predictions <- predict(imm_biomass, newdata = nd)
    saveRDS(imm_predictions, file = paste0("data/", spp, 
      "/predictions-", spp, covs, "-", ssid_string, "-imm-biomass-prior-", priors, ".rds"
    ))
  } else {
    rm(adult_predictions)
    rm(imm_predictions)

@@

      total_biomass <- readRDS(paste0("data/", spp, 
        "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
      ))
      predictions <- predict(total_biomass, newdata = nd)
      saveRDS(predictions, file = paste0("data/", spp, 
        "/predictions-", spp, covs, "-", ssid_string, "-total-biomass-prior-", priors, ".rds"
      ))
    }
  }
```

>>>>

Species
```{r}

```

@@

knitr::opts_chunk$set(
  fig.width = 11, fig.height = 8.5,
  fig.path = paste0("figs/", spp, "/"),
  echo = FALSE, warning = FALSE, message = FALSE
)
```

@@

# scale predictors before filtering to ensure mean and SD are global
data <- data %>% mutate(raw_depth = depth, depth = log(raw_depth))
data <- gfranges::scale_predictors(data, # predictors = c(quo(depth)))
  predictors = c(quo(depth), quo(mixed), quo(muddy), quo(sandy), quo(rocky), quo(any_rock))
)
data <- data %>%
  mutate(depth = raw_depth) %>%
  filter(ssid %in% model_ssid)
```

@@

Run climate independent sdmTMB model
```{r}
tictoc::tic()
if (any(names(data) == "adult_density")) {
  adult_formula <- as.formula(paste(
    "adult_density ~ 0 + as.factor(year)", covariates, ""
  ))

  starttime1 <- Sys.time()
  adult_biomass <- sdmTMB::sdmTMB(data,
    adult_formula,
    time_varying = ~ 0 + depth_scaled + depth_scaled2, #+ trawled
    time = "year", spde = spde,
    family = tweedie(link = "log"),
    ar1_fields = TRUE,
    include_spatial = TRUE,
    enable_priors = priors,
    control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
    silent = FALSE
  )
  endtime1 <- Sys.time()
  time1 <- round(starttime1 - endtime1)

  saveRDS(adult_biomass,
    file = paste0(
      "data/", spp,
      "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    )
  )

@@

  saveRDS(imm_biomass,
    file = paste0(
      "data/", spp,
      "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    )
  )
} else {
  starttime3 <- Sys.time()
  dens_formula <- as.formula(paste("density ~ 0 + as.factor(year)", covariates, ""))

  total_biomass <- sdmTMB::sdmTMB(data,
    dens_formula,
    time_varying = ~ 0 + depth_scaled + depth_scaled2, # + trawled
    time = "year", spde = spde,
    family = tweedie(link = "log"),
    ar1_fields = TRUE,
    include_spatial = TRUE,
    enable_priors = priors,
    control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
    silent = FALSE
  )
  endtime3 <- Sys.time()
  time3 <- round(starttime3 - endtime3)

  saveRDS(total_biomass, file = paste0(
    "data/", spp,
    "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))
}
tictoc::toc()
```

@@

  bath <- readRDS("data/bathymetry-data")
  # dat <- bath$data %>% scale_survey_predictors()
  dat <- bath$data %>% mutate(raw_depth = depth, depth = log(raw_depth))
  dat <- gfranges::scale_predictors(dat, predictors = c(quo(depth)))

@@

  newcov <- readRDS("data/new_covariates.rds") %>% dplyr::select(X, Y, trawled, muddy, rocky, mixed, any_rock)
  nd_all <- left_join(nd_all, newcov, by = c("X", "Y"))

@@

```{r}
# covs <- gsub("\\+", "-", covariates)
fish <- readRDS(paste0("raw/bio-data-", spp, ""))

@@

if (length(unique(fish$maturity_code)) > 2) {
  nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
  adult_biomass <- readRDS(paste0(
    "data/", spp,
    "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))

@@

  adult_predictions <- predict(adult_biomass, newdata = nd)
  saveRDS(adult_predictions, file = paste0(
    "data/", spp,
    "/predictions-", spp, covs, "-", ssid_string, "-mature-biomass-prior-", priors, ".rds"
  ))

  imm_biomass <- readRDS(paste0(
    "data/", spp,
    "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
  ))
  imm_predictions <- predict(imm_biomass, newdata = nd)
  saveRDS(imm_predictions, file = paste0(
    "data/", spp,
    "/predictions-", spp, covs, "-", ssid_string, "-imm-biomass-prior-", priors, ".rds"
  ))
} else {
  rm(adult_predictions)
  rm(imm_predictions)

@@

    total_biomass <- readRDS(paste0(
      "data/", spp,
      "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    ))
    predictions <- predict(total_biomass, newdata = nd)
    saveRDS(predictions, file = paste0(
      "data/", spp,
      "/predictions-", spp, covs, "-", ssid_string, "-total-biomass-prior-", priors, ".rds"
    ))
  }
}
```
